\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-26 misc/JSS article/final/article.tex
Tweaks and hyphenation.

@@

R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)

@@

Using the \emph{Fast Statistical Functions} directly often results in more efficient code. For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than \code{POP / fsum(POP)} (extra allocation) or \code{proportions(POP)} (split-apply-combine). \newline

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing, but all options can be distinguished in efficiency terms. It is thus useful to be mindful of computer resources and attempt to craft minimalistic solutions. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type. This is very efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}.

@@

\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, implying that using any \emph{Fast Statistical Function} in an expression causes the whole expression to be vectorized. This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated split-apply-combine. The eager vectorization approach enables efficient grouped calculation of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.

@@

When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to \proglang{C/C++}, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)}, amounting to \code{cov(x, y)/var(y)}, is thus vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, running 1.6 million cell-regressions and obtaining 2 forecasts for 2021 and 2022 took only 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline

@@

1   collapse   3.32µs   4.18µs   14.33KB  9999    1    43.08ms
2      dplyr    200µs 238.91µs    1.53MB  9934   66      2.48s

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. In the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
\begin{Schunk}
\begin{Sinput}
R> exportsi <- exports |> findex_by(c, s, y)
R> exportsi |> G() |> head(5)
\end{Sinput}
\begin{Soutput}
   c  s    y   G1.v
1 c1 s1 2002     NA
2 c1 s1 2003  22.17
3 c1 s1 2004 -68.24
4 c1 s1 2005  44.98
5 c1 s1 2006 277.76

@@

c.s [100] | y [10]
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
\begin{Schunk}
\begin{Sinput}
R> vi <- exportsi$v; str(vi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:500] 0.552 0.675 0.214 0.311 1.174 ...
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ c.s: Factor w/ 100 levels "c1.s1","c2.s1",..: 1 1 1 1 1 1 1 1 ..
  ..$ y  : Ord.factor w/ 10 levels "2001"<"2002"<..: 2 3 4 5 6 7 8 9..
  ..- attr(*, "nam")= chr [1:3] "c" "s" "y"
\end{Soutput}
\begin{Sinput}
R> is_irregular(vi)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> vi |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001  2002  2003  2004  2005 2006  2007   2008  2009  2010
c1.s1   NA 0.552 0.675 0.214 0.311 1.17 0.619 0.1127 0.917 0.223
c2.s1   NA 0.795    NA    NA 0.237   NA    NA 0.0585 0.818    NA
c3.s1   NA 0.709 0.268 1.464    NA   NA 0.467 0.1193 0.467    NA
\end{Soutput}
\begin{Sinput}
R> fdiff(vi) |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001 2002   2003   2004   2005  2006   2007   2008  2009   2010
c1.s1   NA   NA  0.122 -0.461 0.0964 0.863 -0.555 -0.506 0.804 -0.694
c2.s1   NA   NA     NA     NA     NA    NA     NA     NA 0.759     NA
c3.s1   NA   NA -0.441  1.196     NA    NA     NA -0.348 0.348     NA
\end{Soutput}
\end{Schunk}
\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
\begin{Schunk}
\begin{Sinput}
R> settransform(exportsi, v_ld = Dlog(v))
R> lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
\end{Sinput}
\begin{Soutput}
               Estimate Std. Error t value Pr(>|t|)
(Intercept)      -0.008      0.141  -0.058    0.954
L(v_ld, 1:2)L1   -0.349      0.115  -3.042    0.004
L(v_ld, 1:2)L2   -0.033      0.154  -0.215    0.831
\end{Soutput}
\end{Schunk}
Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions. An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
%
\begin{Schunk}
\begin{Sinput}
R> BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
\end{Sinput}
\begin{Soutput}
 [1]     NA     NA     NA     NA 0.5853 0.5986 0.4861 0.6267 0.6092     NA

@@

\fct{join} is thus a highly efficient, versatile, and verbose implementation of table joins for \proglang{R}.
%
\subsection{Pivots}
%
The reshaping/pivoting functionality of both commercial and open-source software is also presently unsatisfactory for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different reshaping functions, such as \fct{data.table::melt}/\fct{tidyr::pivot\_longer} to combine columns and \fct{data.table::dcast}/\fct{tidyr::pivot\_wider} to expand them, requiring users to learn both. Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two functions. \newline

\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+                   labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

R> qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)

@@

High income          3828  5.87959699e+10  75.6926  4.5346  42.672  85.4171
Low income           1792  2.09491614e+10  53.5061  8.8691  26.172    74.43
Lower middle income  2790  1.13837685e+11  60.5865  8.3607  18.907   76.699
Upper middle income  3249  1.19295269e+11  68.2698  7.1889  36.535   80.279

@@

R> wlda15 |> descr(by = income + LIFEEX ~ OECD)

@@

1     base_int 43.78ms  45.3ms   166.2MB    65   65         3s
2 collapse_int  6.25ms  8.24ms    38.2MB   353   71         3s

@@

1     base_char 69.9ms 71.3ms   166.2MB    43   43      3.07s
2 collapse_char 13.9ms 15.6ms    38.2MB   180   36      3.01s

@@

1     base_int 16.92ms  17.4ms    76.3MB    78   78      1.37s
2 collapse_int  6.95ms  7.68ms    38.2MB   270   75      2.08s

@@

1       base_char 58.8ms 58.8ms   114.5MB     1   48    58.78ms
2 data.table_char 32.2ms 33.7ms    38.1MB    66   19      2.22s
3   collapse_char 19.5ms 20.4ms    38.1MB   113   28      2.31s

@@

1          R   5.4ms   6.2ms    7.86KB   493    0         3s
2   collapse 282.9µs 324.5µs    7.86KB  9150    1         3s

@@

1          R   5.38ms   6.2ms   32.09KB   491    0         3s
2   collapse 280.93µs 327.2µs    7.86KB  8975    0         3s

@@

1         MS 77.8ms 79.4ms    89.9KB    38    0      3.02s
2   collapse 19.1ms 19.5ms    27.2KB   154    0      3.01s

@@

1          R   5.4ms 5.62ms    7.87MB   459   35      2.61s
2   collapse 900.2µs 1.13ms    7.71MB  1876  104      2.19s

@@

1      dplyr 363.83ms 479.56ms    50.5MB     7   65      3.15s
2 data.table   7.22ms   7.99ms    20.8MB   339   33         3s
3   collapse   2.96ms    3.4ms     9.2MB   769   33         3s

@@

1      dplyr_mean  1.16s  1.17s   48.57MB     3   78      3.52s
2 data.table_mean 7.43ms 8.31ms   18.93MB   283   28         3s
3   collapse_mean  3.1ms 3.71ms    9.11MB   679   28         3s

@@

1      dplyr_median    4.2s    4.2s    52.8MB     1   91       4.2s

>>>>

% Hypthens
\hyphenation{mod-i-fi-ca-tions ma-nip-u-la-tion trans-for-ma-tion per-for-mance de-vel-op-ment
             ef-fi-cient in-tro-duces ca-pa-bil-i-ties gen-er-al-iz-ing en-abling pro-gram-ming
             com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries com-pu-ta-tion-al}
\usepackage[english]{babel}
% \sloppy
% \usepackage{microtype}

@@

R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(g = iris$Species)

@@

Emplyoing the \emph{Fast Statistical Functions} directly often results in more efficient code. For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(\\POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than using \code{POP / fsum(POP)} (extra full-length allocation) or \code{proportions(POP)} (slow split-apply-combine evaluation logic) inside the grouped \fct{fmutate} call. \newline

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing at first, but all options can be distinguished in terms of efficiency. It is thus helpful to be mindful of computer resources and attempt to craft minimalistic solutions to complex problems. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type/category. After creating comparable POI confidence, location, and type indicators, my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (richest) source by location and type. This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} internal grouped logic is highly optimized.

@@

\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Function} returning grouped output). This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated using split-apply-combine logic. This eager vectorization approach enables efficient grouped calculations of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.

@@

When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to \proglang{C/C++}, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, I ran 1.6 million cell-regressions and obtained 2 forecasts for 2021 and 2022 in less than 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline

@@

1   collapse   3.81µs   4.22µs        0B  2235    0     9.91ms
2      dplyr 239.44µs 275.77µs    1.92KB    33    0    10.35ms

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 44.7 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

An example using the \code{exports} data follows. \emph{Note} that data need not be sorted for indexing.
\begin{Schunk}
\begin{Sinput}
R> exportsi <- exports |> findex_by(c, s, y)
R> exportsi |> G() |> head(5)
\end{Sinput}
\begin{Soutput}
   c  s    y   G1.v
1 c1 s1 2002     NA
2 c1 s1 2003  22.17
3 c1 s1 2004 -68.24
4 c1 s1 2005  44.98
5 c1 s1 2006 277.76

@@

c.s [100] | y [10]
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
\begin{Schunk}
\begin{Sinput}
R> vi <- exportsi$v; str(vi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:500] 0.552 0.675 0.214 0.311 1.174 ...
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ c.s: Factor w/ 100 levels "c1.s1","c2.s1",..: 1 1 1 1 1 1 1 1 ..
  ..$ y  : Ord.factor w/ 10 levels "2001"<"2002"<..: 2 3 4 5 6 7 8 9..
  ..- attr(*, "nam")= chr [1:3] "c" "s" "y"
\end{Soutput}
\begin{Sinput}
R> is_irregular(vi)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> vi |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001  2002  2003  2004  2005 2006  2007   2008  2009  2010
c1.s1   NA 0.552 0.675 0.214 0.311 1.17 0.619 0.1127 0.917 0.223
c2.s1   NA 0.795    NA    NA 0.237   NA    NA 0.0585 0.818    NA
c3.s1   NA 0.709 0.268 1.464    NA   NA 0.467 0.1193 0.467    NA
\end{Soutput}
\begin{Sinput}
R> fdiff(vi) |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001 2002   2003   2004   2005  2006   2007   2008  2009   2010
c1.s1   NA   NA  0.122 -0.461 0.0964 0.863 -0.555 -0.506 0.804 -0.694
c2.s1   NA   NA     NA     NA     NA    NA     NA     NA 0.759     NA
c3.s1   NA   NA -0.441  1.196     NA    NA     NA -0.348 0.348     NA
\end{Soutput}
\end{Schunk}
\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
\begin{Schunk}
\begin{Sinput}
R> settransform(exportsi, v_ld = Dlog(v))
R> lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
\end{Sinput}
\begin{Soutput}
               Estimate Std. Error t value Pr(>|t|)
(Intercept)      -0.008      0.141  -0.058    0.954
L(v_ld, 1:2)L1   -0.349      0.115  -3.042    0.004
L(v_ld, 1:2)L2   -0.033      0.154  -0.215    0.831
\end{Soutput}
\end{Schunk}
Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions. For example, below I compute a 5-year rolling average (\fct{ix} abbreviates \fct{findex}).
%
\begin{Schunk}
\begin{Sinput}
R> BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
\end{Sinput}
\begin{Soutput}
 [1]     NA     NA     NA     NA 0.5853 0.5986 0.4861 0.6267 0.6092     NA

@@

\fct{join} is thus a highly efficient, versatile, and verbose implementation of table joins for \proglang{R}.
%
\subsection{Pivots}
%
The reshaping/pivoting functionality of both commercial and open-source software is also presently unsatisfactory for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can presently only be reshaped by losing these labels or manual efforts to retain them. Modern \proglang{R} packages also offer different reshaping functions, such as \fct{data.table::melt}/\fct{tidyr::pivot\_longer} to combine columns and \fct{data.table::dcast}/\fct{tidyr::pivot\_wider} to expand them, requiring users to learn both. Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two functions. \newline

\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+   fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+   fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+    labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. A small number extremely fast internal aggregation functions, \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"} operate 'on the fly' while reshaping. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided by the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

R> qsu(wlda15, by = LIFEEX ~ income, w = ~ POP)

@@

High income          68  1.19122607e+09   80.879   2.441  70.6224  85.078
Low income           29      694'893773  63.8061  3.9266   53.283  72.697
Lower middle income  47  3.06353648e+09  68.7599  4.7055   54.331  76.699
Upper middle income  55  2.67050662e+09  75.9476  2.3895   58.735  80.279

@@

R> descr(wlda15, by = income + LIFEEX ~ OECD)

@@

1     base_int 46.5ms 46.5ms   166.2MB     1    0     46.5ms
2 collapse_int 10.5ms 10.5ms    38.2MB     1    0     10.5ms

@@

1     base_char 69.8ms 69.8ms   166.2MB     1    0     69.8ms
2 collapse_char 20.6ms 20.6ms    38.2MB     1    0     20.6ms

@@

1     base_int 27.28ms 27.28ms    76.3MB     1    1     27.3ms
2 collapse_int  7.93ms  8.01ms    38.2MB     2    0       16ms

@@

1       base_char 72.4ms 72.4ms   114.5MB     1    0     72.4ms
2 data.table_char 46.5ms 46.5ms    38.1MB     1    1     46.5ms
3   collapse_char   29ms   29ms    38.1MB     1    0       29ms

@@

1          R   6.07ms   6.09ms    7.86KB     2    0     12.2ms
2   collapse 311.31µs 345.01µs    7.86KB    28    0     10.1ms

@@

1          R   6.22ms   6.22ms    7.86KB     2    0     12.4ms
2   collapse 329.27µs 341.08µs    7.86KB    27    0     10.3ms

@@

1         MS 77.5ms 77.5ms   86.04KB     1    0     77.5ms
2   collapse 19.5ms 19.5ms    7.86KB     1    0     19.5ms

@@

1          R 6.04ms 6.29ms    7.85MB     2    0     12.6ms
2   collapse  1.3ms 1.39ms    7.67MB     8    0     11.2ms

@@

1      dplyr 369.09ms 369.09ms   48.56MB     1    6    369.1ms
2 data.table   8.36ms   8.46ms   18.93MB     2    0     16.9ms
3   collapse   4.24ms   4.29ms    9.11MB     3    0     12.9ms

@@

1      dplyr_mean  1.28s  1.28s   48.56MB     1   23      1.28s
2 data.table_mean  8.9ms 9.05ms   18.93MB     2    0    18.11ms
3   collapse_mean 4.08ms 4.39ms    9.11MB     3    0    13.02ms

@@

1      dplyr_median   4.51s   4.51s    52.8MB     1   90      4.51s
